Deep Learning does not Replace Bayesian Modeling

### 2021-08-26 DeepLearningAndBayesianSoftware.Rmd
submitted to journal

---
title: "Deep Learning and Bayesian Modeling"
subtitle: "Comparing research use via citation counting"
author: "Breck Baldwin"
email: "breckbaldwin@gmail.com"
date: "5/23/2021"
output:
 bookdown::html_document2:
   includes:
     in_header: _html/ga.html 
bibliography: citations.bib
link-citations: yes
---

@@

Stan, a Bayesian modeling language, was released in 2012 to considerable fanfare. A shiny new inference algorithm, HMC with NUTS (@hoffman2014no), promised and delivered fits that could not be fit before. And then deep learning happened. The release of TensorFlow (@abadi2016tensorflow) opened the doors of deep learning to everyone and in 2016 began getting significant traction. One result of this was people often assumed that deep learning's successes usurped Bayesian modeling's domain. This was not in our imagination, NSF reviews came back dismissing Stan funding because all the interesting work was assumed to be happening with deep learning. Mind you, deep learning and Bayesian modeling are conceptual cousins but very different from each other and are better thought of as complementary than as antagonistic. Yet Bayesians found themselves in deep learning's shadow somehow.

@@

1. How does Bayesian modeling software stack up against deep learning? 
2. Asses the impact of Bayesian modeling software using deep learning metrics as a yard stick--how big a fraction of a huge thing are we?
3. Contextualize the roles each approach has by looking at subject distributions. The technologies have very different use cases so one would expect variation.

@@

[^1]:There are also lighter weight interfaces as well interface to other languages that I ignore due to low expected research mentions, see [https://https://mc-stan.org/users/interfaces/]([https://https://mc-stan.org/users/interfaces/) for a listing.

The key resource behind this document is Elsevier's https://scopus.com research search engine that provides a tightly curated[^2] search index that includes sources outside of the academic behemoth's own journals. It also provides a solid API (Application Programmer Interface) and a classification of journals into subject areas. The actual form of Scopus queries is discussed below.

[^2]:https://scolar.google.com genrally yields double the counts for similar queries but no subject classification or API to code against. The increased counts are likely due to inclusion of non-peer reviewed sites like [https://arxiv.org](https://arxiv.org)

@@

For an example of how the queries were developed, the stand alone query 'rstan' had the addition of rejecting articles that mentioned 'mit' because there was a common reference to 'http://wwwmath.mit.edu/~rstan/ec/' that existed prior to 2012 when Stan was released. The name 'Stan' presented obvious difficulties, see [https://statmodeling.stat.columbia.edu/2019/04/29/we-shouldntve-called-it-stan-i-shouldve-listened-to-bob-and-hadley/](https://statmodeling.stat.columbia.edu/2019/04/29/we-shouldntve-called-it-stan-i-shouldve-listened-to-bob-and-hadley/). The query shown achieved desired count drop off at 2012 with reasonable precision and each of the 'OR' terms contributed to coverage significantly. How the word 'mc-stan.org' was tokenized is unknown but there were no results before 2013 for the solo query.

@@

The above queries can be run with the advance search option at https://scopus.com and are generated from the API queries actually used. The change is replacing '+' with ' '. An account is required which is included for many education institutions.

@@

I am waiting on some feedback from the Stan community before making any observations. But I will say that this overall has me feeling quite optimistic about the future of Bayesian modeling.

@@

I intend this to be an evolving document with periodic updates at it's github repo at [https://github.com/breckbaldwin/ScientificSoftwareImpactMetrics](https://github.com/breckbaldwin/ScientificSoftwareImpactMetrics) where it lives as `DeepLearningAndBayesianSoftware.Rmd`, is rendered as `DeepLearningAndBayesianSoftware.html` and is viewable as [https://breckbaldwin.github.io/ScientificSoftwareImpactMetrics/DeepLearningAndBayesianSoftware.html](https://breckbaldwin.github.io/ScientificSoftwareImpactMetrics/DeepLearningAndBayesianSoftware.html). Note that github pages changes urls around in odd ways for the rendered view of hte html page. All source code is viewable in the .Rmd document.

>>>>

---
title: "Deep Learning does not Replace Bayesian Modeling"
subtitle: "Comparing research use via citation counting"
author: "Breck Baldwin"
email: "breckbaldwin@gmail.com"
date: "8/20/2021"
output:
 bookdown::html_document2:
   includes:
     in_header: _html/ga.html

bibliography: citations.bib
csl: american-medical-association.csl
link-citations: yes
---

@@

Stan, a Bayesian modeling language, was released in 2012 to considerable fanfare. A shiny new inference algorithm, HMC with NUTS (@hoffman2014no), promised and delivered fits that could not be fit before. And then deep learning happened. The release of TensorFlow (@abadi2016tensorflow) opened the doors of deep learning to all and by it 2016 began getting significant traction. One result of this was people often assumed that deep learning's successes usurped Bayesian modeling's domain. This is not in our collective imagination--Bayesians like to 'believe things' after all--NSF reviews came back dismissing Stan funding because all the interesting work was assumed to be happening with deep learning. Recently in the UK, open skepticism was expressed about impact claims for Bayesian software in response to a research grant. Mind you, deep learning and Bayesian modeling are conceptual cousins but in the end are very different from each other. They are better thought of as complementary than as antagonistic. Yet Bayesians found themselves in deep learning's shadow somehow.

@@

1. How does Bayesian modeling software stack up against deep learning without appeal to feature comparisons, performance arguments on suspect data sets or achieved closeness to Platonic ideals? Just go out and count how many citations the respective approaches have in the research literature. Counting and categorization, that's it. 
2. Asses the impact of Bayesian modeling software using deep learning metrics as a yard stick--how big a fraction of a huge thing are we?
3. Contextualize the roles each approach has by looking at subject distributions. The technologies have very different use cases so one would expect variation.

@@

[^1]:There are also lighter weight interfaces as well interface to other languages that I ignore due to low expected research mentions, see https://https://mc-stan.org/users/interfaces/ for a listing.

The key resource behind this document is Elsevier's https://scopus.com research search engine that provides a tightly curated[^2] search index that includes sources outside of the academic behemoth's own journals. It also provides a solid API (Application Programmer Interface) and a classification of journals into subject areas. The actual form of Scopus queries is discussed below which should allow the questioning reader to verify and update the counts for the various packages discussed.

[^2]:https://scolar.google.com genrally yields double the counts for similar queries but no subject classification or API to code against. The increased counts are likely due to inclusion of non-peer reviewed sites like https://arxiv.org

<br>

@@

For an example of how the queries were developed, the stand alone query 'rstan' had the addition of rejecting articles that mentioned 'mit' because there was a common reference to 'http://wwwmath.mit.edu/~rstan/ec/' that existed prior to 2012 when Stan was released. The name 'Stan' presented obvious difficulties, see https://statmodeling.stat.columbia.edu/2019/04/29/we-shouldntve-called-it-stan-i-shouldve-listened-to-bob-and-hadley/. The query shown achieved desired count drop off at 2012 with reasonable precision and each of the 'OR' terms contributed to coverage significantly. How the word 'mc-stan.org' was tokenized is unknown but there were no results before 2013 for the solo query.

@@

The above queries can be run with the advance search option at https://scopus.com and are generated from the API queries actually used where '+' was replaced with ' ' above. An account is required which is included for many education institutions.

@@

As often happens conducting a closer look reveals subtleties. Deep learning dominates the computer science part of science and obviously it dominates in the commercial sector which this article does not address. However the remaining parts of science show a more balanced distribution across the approaches. This makes sense since the approaches are very different and serve different goals.

As a Bayesian advocate I'd like to leave the funding agencies with the observation that Bayesian software carries roughly one third of the research load by research citation count (3925/(3925+8884)) without the backing of Fortune 500 companies like Facebook and Google. Just recently I learned of a funding denial because the funders didn't believe that Bayesian software could have the impact that was being claimed--that opinion could well be self fulfilling in the future but as of now it is not justified.

@@

I intend this to be an evolving document with periodic updates at it's github repo at https://github.com/breckbaldwin/ScientificSoftwareImpactMetrics where it lives as `DeepLearningAndBayesianSoftware.Rmd`, is rendered as `DeepLearningAndBayesianSoftware.html` and is viewable as https://breckbaldwin.github.io/ScientificSoftwareImpactMetrics/DeepLearningAndBayesianSoftware.html. Note that github pages changes urls around in odd ways for the rendered view of hte html page. All source code is viewable in the .Rmd document.
